---
title: "Lab 11"
output: html_document
author: Alina Sequeira
date: 12-3-20
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dada2 Pipeline Tutorial
```{r}
if(!requireNamespace("BiocManager")){
  install.packages("BiocManager")
}
BiocManager::install("phyloseq")
```
```{r}
BiocManager::install("dada2")
```
```{r}
BiocManager::install("DECIPHER")
```
```{r}
library(BiocManager)
```
```{r}
library(phyloseq)
```
```{r}
library(DECIPHER)
```

```{r}
library(dada2); packageVersion("dada2")
```

```{r}
path <- "data/MiSeq_SOP"
list.files(path)
```
```{r}
# Forward and reverse fastq filenames have format: SAMPLENAME_R1_001.fastq and SAMPLENAME_R2_001.fastq
fnFs <- sort(list.files(path, pattern="_R1_001.fastq", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_R2_001.fastq", full.names = TRUE))

# Extract sample names, assuming filenames have format: SAMPLENAME_XXX.fastq
sample.names <- sapply(strsplit(basename(fnFs), "_"), `[`, 1)
```

## Inspect Read Quality Profiles
Visualizing quality profiles of forward reads
```{r}
plotQualityProfile(fnFs[1:2])
```
```{r}
# mean quality score at each position is shown by the green line
# the quartiles of the quality score distribution by the orange lines
# red line shows the scaled proportion of reads that extend to at least that position 
# generally advise trimming the last few nucleotides to avoid less well-controlled errors that can arise there (not needed here)
```

Visualing quality profiles of reverse reads
```{r}
plotQualityProfile(fnRs[1:2])
```

```{r}
# reverse reads are of significantly worse quality, especially at the end
# trimming as the average qualities crash will improve the algorithm’s sensitivity to rare sequence variants
# Based on these profiles, we will truncate the reverse reads at position 160 where the quality distribution crashes
```

## Filter and Trim
```{r}
# Assign the filenames for the filtered fastq.gz files.

# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "filtered", paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names
```

```{r}
#  use standard filtering parameters: maxN=0 (DADA2 requires no Ns), truncQ=2, rm.phix=TRUE and maxEE=2
#  maxEE parameter sets the maximum number of “expected errors” allowed in a read

out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(240,160),
              maxN=0, maxEE=c(2,2), truncQ=2, rm.phix=TRUE,
              compress=TRUE, multithread=TRUE) # On Windows set multithread=FALSE
```

```{r}
head(out)
```
Considerations for your own data: The standard filtering parameters are starting points, not set in stone. If you want to speed up downstream computation, consider tightening maxEE. If too few reads are passing the filter, consider relaxing maxEE, perhaps especially on the reverse reads (eg. maxEE=c(2,5)), and reducing the truncLen to remove low quality tails. Remember though, when choosing truncLen for paired-end reads you must maintain overlap after truncation in order to merge them later.

Considerations for your own data: For ITS sequencing, it is usually undesirable to truncate reads to a fixed length due to the large length variation at that locus. That is OK, just leave out truncLen. See the DADA2 ITS workflow for more information.

## Learning the Error Rates
* DADA2 algorithm makes use of a parametric error model (err) and every amplicon dataset has a different set of error rates
* learnErrors method learns this error model from the data, by alternating estimation of the error rates and inference of sample composition until they converge on a jointly consistent solution
```{r}
errF <- learnErrors(filtFs, multithread=TRUE)
```
```{r}
errR <- learnErrors(filtRs, multithread=TRUE)
```
```{r}
# Visualize the estimated error rates
plotErrors(errF, nominalQ=TRUE)
```
```{r}
# The error rates for each possible transition (A→C, A→G, …) are shown
# Points are the observed error rates for each consensus quality score
# Black line shows the estimated error rates after convergence of the machine-learning algorithm
# Red line shows the error rates expected under the nominal definition of the Q-score
# Here the estimated error rates (black line) are a good fit to the observed rates (points), and the error rates drop with increased quality as expected, CAN PROCEED W/ CONFIDENCE 
```

## Sample Interface
```{r}
# apply the core sample interference algorithm to the filtered and trimmed sequence data

dadaFs <- dada(filtFs, err=errF, multithread=TRUE)
```
```{r}
dadaRs <- dada(filtRs, err=errR, multithread=TRUE)
```
```{r}
# inspecting the returned data-class object
dadaFs[[1]]
```
```{r}
# The DADA2 algorithm inferred 128 true sequence variants from the 1979 unique sequences in the first sample. There is much more to the dada-class return object than this (see help("dada-class") for some info), including multiple diagnostics about the quality of each denoised sequence variant
```

## Merge Paired Reads
* Merge the forward and reverse reads together to obtain the full denoised sequences
* Merging is performed by aligning the denoised forward reads with the reverse-complement of the corresponding denoised reverse reads, and then constructing the merged “contig” sequences
*By default, merged sequences are only output if the forward and reverse reads overlap by at least 12 bases, and are identical to each other in the overlap region (but these conditions can be changed via function arguments)
```{r}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)
```
```{r}
# Inspect the merger data.frame from the first sample
head(mergers[[1]])
```
## Constrcut Sequence Table
```{r}
# construct an amplicon sequence variant table (ASV) table, a higher-resolution version of the OTU table produced by traditional methods.

seqtab <- makeSequenceTable(mergers)
dim(seqtab)
```
```{r}
# Inspect distribution of sequence lengths

table(nchar(getSequences(seqtab)))
```
```{r}
# The sequence table is a matrix with rows corresponding to (and named by) the samples, and columns corresponding to (and named by) the sequence variants
# This table contains 293 ASVs, and the lengths of our merged sequences all fall within the expected range for this V4 amplicon.
```

## Remove Chimeras
* The core dada method corrects substitution and indel errors, but chimeras remain.
* Chimeric sequences are identified if they can be exactly reconstructed by combining a left-segment and a right-segment from two more abundant “parent” sequences
```{r}
seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)
```
```{r}
dim(seqtab.nochim)
```
```{r}
sum(seqtab.nochim)/sum(seqtab)
```
```{r}
# The frequency of chimeric sequences varies substantially from dataset to dataset, and depends on on factors including experimental procedures and sample complexity. Here chimeras make up about 21% of the merged sequence variants, but when we account for the abundances of those variants we see they account for only about 4% of the merged sequence reads.
```

Considerations for your own data: Most of your reads should remain after chimera removal (it is not uncommon for a majority of sequence variants to be removed though). If most of your reads were removed as chimeric, upstream processing may need to be revisited. In almost all cases this is caused by primer sequences with ambiguous nucleotides that were not removed prior to beginning the DADA2 pipeline

## Track Reads Through the Pipeline
```{r}
# As a final check of our progress, we’ll look at the number of reads that made it through each step in the pipeline:

getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim))

# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)

colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
head(track)
```
```{r}
# Looks good! We kept the majority of our raw reads, and there is no over-large drop associated with any single step.
```

Considerations for your own data: This is a great place to do a last sanity check. Outside of filtering, there should no step in which a majority of reads are lost. If a majority of reads failed to merge, you may need to revisit the truncLen parameter used in the filtering step and make sure that the truncated reads span your amplicon. If a majority of reads were removed as chimeric, you may need to revisit the removal of primers, as the ambiguous nucleotides in unremoved primers interfere with chimera identification

## Assign Taxonomy
The assignTaxonomy function takes as input a set of sequences to be classified and a training set of reference sequences with known taxonomy, and outputs taxonomic assignments with at least minBoot bootstrap confidence
```{r}
# Unsurprisingly, the Bacteroidetes are well represented among the most abundant taxa in these fecal samples. Few species assignments were made, both because it is often not possible to make unambiguous species assignments from subsegments of the 16S gene, and because there is surprisingly little coverage of the indigenous mouse gut microbiota in reference databases.
```
Considerations for your own data: If your reads do not seem to be appropriately assigned, for example lots of your bacterial 16S sequences are being assigned as Eukaryota NA NA NA NA NA, your reads may be in the opposite orientation as the reference database. Tell dada2 to try the reverse-complement orientation with assignTaxonomy(…, tryRC=TRUE) and see if this fixes the assignments. If using DECIPHER for taxonomy, try IdTaxa (…, strand=“both”).

## Evaluating Accuracy
```{r}
# Evaluating DADA2's accuracy on the mock community:

unqs.mock <- seqtab.nochim["Mock",]
unqs.mock <- sort(unqs.mock[unqs.mock>0], decreasing=TRUE) # Drop ASVs absent in the Mock
cat("DADA2 inferred", length(unqs.mock), "sample sequences present in the Mock community.\n")
```
```{r}
mock.ref <- getSequences(file.path(path, "HMP_MOCK.v35.fasta"))
match.ref <- sum(sapply(names(unqs.mock), function(x) any(grepl(x, mock.ref))))
cat("Of those,", sum(match.ref), "were exact matches to the expected reference sequences.\n")
```

